Pan-cancer analysis

### 2021-03-11 src/manuscript_figures.Rmd
update figures; reorganize, keep figure counters, change figure filenames

# todo: move this to settings file
inputDir <- '/data/local/buyar/arcas/pancancer_multiomics_manuscript/output'
assay_dir <- settings$assay_output$folder
surv <- do.call(rbind, readRDS(settings$surv))
projectDict <- sapply(yaml::read_yaml(settings$projectDict), 
                      function(x) gsub(' ', '', unlist(strsplit(x, ','))))
tcga_subtypes <- process_tcga_subtypes()

Figures <- c() # accummulate all figures in this
SuppFigures <- c() # accumulate all supplementary figures in this

```

```{r import_lfs}
LFs <- sapply(simplify = F, c('pca_factors', 'maui_factors', 'mofa_factors'), function(folder) {
  message('Importing ',folder)
  files <- dir(file.path(inputDir, folder), pattern = 'factors.csv$', full.names = T)
  LFs <- pbapply::pbsapply(simplify = F, files, function(f) {
    dt <- data.table::fread(f)
    M <- as.matrix(dt[,-1])
    rownames(M) <- dt$V1
    return(M)
  })
  names(LFs) <- unlist(lapply(strsplit(basename(names(LFs)), split = '\\.'), function(x) paste(x[1:2], collapse = '.')))
  return(LFs)
})
```

```{r import_feature_weights}
FWs <- sapply(simplify = F, c('pca_factors', 'maui_factors', 'mofa_factors'), function(folder) {
  message('Importing ',folder)
  files <- dir(file.path(inputDir, folder), pattern = 'feature_weights.csv$', full.names = T)
  W <- pbapply::pbsapply(simplify = F, files, function(f) {
    dt <- data.table::fread(f)
    M <- as.matrix(dt[,-1])
    rownames(M) <- gsub("^assay: ", "", dt$V1) # required only for maui
    return(M)
  })
  names(W) <- unlist(lapply(strsplit(basename(names(W)), split = '\\.'), 
                            function(x) paste(x[1:2], collapse = '.')))
  return(W)
})
```

```{r}
# Define a function that takes as input LFs and some patient labels (clinical/cancer type/subtype)
# and evaluates predictability of the labels using the LFs
predict_labels <- function(LFs, surv, tcga_subtypes, 
                           project, datatype, label.type, ...) {
  source('/data/local/buyar/arcas/subtyping_paper/src/common_functions.R')
  predict_stats <- do.call(rbind, sapply(simplify = F, names(LFs), function(tool) {
    require(data.table)
    message("predicting ",label.type," using ",tool)
    experiment <- paste(project, datatype, sep = '.')
    M <- LFs[[tool]][[experiment]]
    if(label.type == 'cancer_type') {
      labels <- data.frame('cancer_type' = surv[match(rownames(M), bcr_patient_barcode)]$project, 
                           row.names = rownames(M))
    } else if (label.type == 'subtype') {
      labels <- data.frame('subtype' = tcga_subtypes[match(rownames(M),
                                                           bcr_patient_barcode)]$Subtype_Selected, 
                           row.names = rownames(M))
    } else if (label.type == 'clinical') {
      labels <- process_clinical_covariates(rownames(M))
    }
    dt <- do.call(rbind, lapply(colnames(labels), function(x) {
      l <- labels[,x]
      if(is.numeric(l)){
        # categorize samples by intervals
        l <- as.character(cut_interval(l, 10))
      }
      l[is.na(l)] <- 'Undefined'
      message("Predicting ",x, " for experiment ",experiment)
      dt <- suppressMessages(predict_cluster_labels(M, l, ...))
      dt$label.type <- x
      dt$project <- project
      dt$datatypes <- datatype
      dt$method <- tool
      return(dt)
    }))
    return(dt)
  }))
  return(predict_stats[!is.na(train_auc)])
}

plot_comp1 <- function(dt) {
  ggboxplot(dt, 
          x = 'method', y = 'test_auc', color = 'method', 
          add = 'jitter', legend = 'none') + 
    stat_compare_means(method = 'wilcox.test', paired = T, method.args = list('alternative' = 'greater'), 
                     comparisons = list(c('maui_factors', 'mofa_factors'), 
                                        c('maui_factors', 'pca_factors'))) +
    labs(y = 'AUC') + theme(axis.text = element_text(size = 10))
}

plot_comp2 <- function(dt) {
  dtc <- melt(dcast(dt, label ~ method, value.var = 'test_auc'), 
              id.vars = c('label', 'maui_factors'))
  dtc$difference <- dtc$maui_factors - dtc$value
  ggplot(dtc, 
         aes(x = maui_factors, y = value)) + 
    geom_point() + geom_abline(slope = 1) + 
    facet_wrap(~ variable, nrow = 1) + 
    labs(y = 'AUC')
}

@@

```{r pancancer_prediction, fig.height=9, fig.width=12}
# Find out if cancer types are predictable using latent factors 
pancancer_stats <- predict_labels(LFs, surv, tcga_subtypes, 'pancancer', 
                                  'cnv_gex_meth_mut', label.type = 'cancer_type', num.trees = 1000)
p1 <- plot_comp1(pancancer_stats) + theme(axis.text.x = element_blank())
p2 <- plot_comp2(pancancer_stats)
# plot cancer specific factors
M <- LFs$maui_factors$pancancer.cnv_gex_meth_mut
ct <- surv[match(rownames(M), bcr_patient_barcode)]$project
p3 <- plot_label_specific_vars(M, ct)

@@

Fig <- plot_grid(plot_grid(p1, p2, nrow = 1, labels = c('A', 'B')), 
                  plot_grid(p3, p4, rel_widths = c(5,3), nrow = 1, labels = c('C', 'D')), 
                  nrow = 2,
                  rel_heights = c(2,3))

Figures <- c(Figures, Fig)
ggsave('pancancer_cancer_type_prediction.pdf', plot = Fig, units = 'in', width = 12, height = 9)

@@

p1 <- plot_comp1(pancancer_subtype_stats) 
p2 <- plot_comp2(pancancer_subtype_stats)
# plot subtype specific factors
M <- LFs$maui_factors$pancancer.cnv_gex_meth_mut
st <- tcga_subtypes[match(rownames(M), bcr_patient_barcode)]$Subtype_Selected
# remove NA values and pick subtypes with at least 20 samples 
selected <- names(which(table(st) > 20))
M <- M[st %in% selected,]
labels <- tcga_subtypes[match(rownames(M), bcr_patient_barcode)]$Subtype_Selected

@@

Fig <- plot_grid(plot_grid(p1, p2, nrow = 2), p3, ncol = 2, rel_widths = c(2, 3), 
          labels = 'AUTO')
print(Fig)

ggsave('pancancer_subtype_prediction.pdf', plot = Fig, units = 'in', width = 12, height = 9)

SuppFigures <- c(SuppFigures, Fig)
```

# Unsupervised pan-cancer clustering

We do unsupervised clustering of samples and measure AMI to observe which tool's clusters correspond best to 
ground truth (cancer types)

```{r unsupervised_clustering, fig.height=12, fig.width=10}
get_plot <- function(M, plot.type = 'cancertype') {

  df <- plot_tsne(M)
  if(plot.type == 'cancertype') {
    df$tcga <- surv[match(rownames(df), bcr_patient_barcode)]$project
    df <- df[!is.na(df$tcga),]
  } else if (plot.type == 'subtype') {
    df$tcga <- tcga_subtypes[match(rownames(df), bcr_patient_barcode)]$Subtype_Selected
    df <- df[!is.na(df$tcga),]
  }
  clust <- get_kmeans_subtypes(M[rownames(df),], k_vals = length(unique(df$tcga)), 0)
  df$cluster <- as.factor(clust[rownames(df),])

  p <- ggplot(df, aes(x = tSNE1, y = tSNE2))  
  p1 <- p + geom_point(aes(color = cluster)) +
    geom_text_repel(data = get_centroid(df[,1:2], df$cluster), 
              aes(x,y,label = text.label), size = 6) + 
    theme(legend.position = 'none')
  p2 <- p + geom_point(aes(color = tcga)) + 
          geom_text_repel(data = get_centroid(df[,1:2], df$tcga), 
              aes(x,y,label = gsub('TCGA-', '', text.label)), size = 3) +
    theme(legend.position = 'none')

  p <- cowplot::plot_grid(p1, p2)
  # also compute mutual information
  ami <- aricode::AMI(df$tcga, df$cluster)
  return(list('p' = p, 'ami' = ami))
}

pr <-  'pancancer' #'pancancer 99%' # 'nsclc 98%' pan-gastrointestinal 98% 'pan-gyn 98%'
e <- paste0(pr, '.cnv_gex_meth_mut')

# for maui, we drop redundant LFs 
p1 <- get_plot(remove_redundant_variables(LFs$maui_factors[[e]], perc = 99), 'cancertype') 
p2 <- get_plot(LFs$pca_factors[[e]], 'cancertype') # use all pcs, 
p3 <- get_plot(LFs$mofa_factors[[e]], 'cancertype') # mofa

# use AMI values to make another plot
p4 <- ggplot(melt(data.frame('MAUI' = p1$ami, 'PCA' = p2$ami, 'MOFA' = p3$ami)), 
       aes(x = variable, y = value)) + 
  geom_bar(aes(fill = variable), stat = 'identity') + 
  geom_text(aes(y = value + 0.05, label = round(value, 2))) + 
  labs(y = 'AMI', x = '') + 
  theme(legend.title = element_blank()) + coord_equal()

Fig <- cowplot::plot_grid(p1$p, p2$p, p3$p, p4, nrow = 4, labels = 'AUTO')
ggsave("pancancer.unsupervised_clustering.pdf", Fig, units = 'in', width = 10, height = 12)

Figures <- c(Figures, Fig)
```

@@

Use top 10 factors (derived from multi-omics) to cluster samples using cancer-specific LFs 
```{r clustering_metrics, fig.width=10, fig.height=8}
multi_omics_comb <- 'cnv_gex_meth_mut'
experiments <- grep(paste0('^TCGA-.*.', multi_omics_comb), names(LFs$pca_factors), value = T)
cl <- parallel::makeCluster(20)
parallel::clusterExport(cl, varlist = c('LFs', 'experiments', 'settings'))
clusters <- sapply(simplify = F, names(LFs), function(tool) {
  message("Clustering top ",tool)
  pbapply::pbsapply(cl = cl, simplify = F, experiments, function(experiment) {
    message(experiment)
    source(settings$utility_script)
    require(data.table)
    M <- LFs[[tool]][[experiment]]
    if(tool == 'maui_factors') {
      Mr <- remove_redundant_variables(M, perc = 99) # !!! make sure to use same setting as used for pan-cancer clustering
      M <- t(subset_features_by_variance(t(Mr), topN = 10))
     } else {
      M <- M[,head(colnames(M), 10),drop = F]
    }
    val <- c('internal') 
    if(ncol(M) > 2) {val <- c('internal', 'stability')}
    res <- clValid::clValid(obj = M,
                          nClust = 3:8,
                          clMethods = 'kmeans',
                          validation = val,
                          maxitems = 10000,
                          nstart = 1000)
    memberships <- as.data.frame(do.call(cbind, lapply(res@clusterObjs[['kmeans']], function(x) x[['cluster']])))
    colnames(memberships) <- paste0('k', colnames(memberships))
    stats <- as.data.table(res@measures)[V1 %in% c('Silhouette', 'APN')]
    colnames(stats) <- c('metric', 'k', 'clustering_method', 'score')
    return(list('clustering_stats' = stats, 'cluster_memberships' = memberships))
  })
})
parallel::stopCluster(cl)

@@

# plot silhouette/stability score comparison
dt1 <- cluster.stats[metric == 'Silhouette'][datatypes == multi_omics_comb]
p1 <- ggboxplot(dt1, 
          x = 'method', y = 'score', add = 'jitter', color = 'method') + 
  stat_compare_means(method = 'wilcox.test', paired = TRUE, 
                     comparisons = list(c('maui_factors', 'pca_factors'), 
                                        c('maui_factors', 'mofa_factors'))) + 
  labs(title = 'Silhouette score', x = '') + 
  theme(legend.position = 'none')

dt2 <- cluster.stats[metric == 'APN'][datatypes == multi_omics_comb]
p2 <- ggboxplot(dt2, 
          x = 'method', y = 'score', add = 'jitter', color = 'method') + 
  stat_compare_means(method = 'wilcox.test', paired = TRUE, 
                     comparisons = list(c('maui_factors', 'pca_factors'), 
                                        c('maui_factors', 'mofa_factors'))) + 
  labs(title = 'Cluster Stability Score (APN)', x = '') + 
  theme(legend.position = 'none')

# combine both scores and plot for different values of k
dt <- dcast(cluster.stats[datatypes == multi_omics_comb], 
            cancer_type + method + k ~ metric, value.var = 'score')
dt$kv <- paste0('kmeans(k=',dt$k,')')
p3 <- ggplot(dt,
       aes(x = Silhouette, y = APN)) + 
        geom_point(aes(color = method), size = 3) + 
        facet_wrap(~ kv, ncol = 2) +
        labs(title = 'Multi-omics clustering: Silhouette vs APN')

Fig <- plot_grid(plot_grid(p1, p2, labels = 'AUTO', ncol = 1),
                  p3, ncol = 2, rel_widths = c(2,3), labels = c('','C'))
ggsave('cluster_stability_metrics.pdf', plot = Fig, units = 'in', 
       width = 10, height = 8)

Figures <- c(Figures, Fig)
```

@@

```{r survival_factors, fig.width=12, fig.height=8}
# factors to use in the model
clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage', 
                  'PFI', 'PFI.time')

experiments <- grep('^TCGA-.*.cnv_gex_meth_mut', names(LFs$maui_factors), value = T)
cl <- parallel::makeCluster(20)
parallel::clusterExport(cl, varlist = c('LFs', 'clin_factors', 
                                        'surv', 'experiments', 'settings'))
surv.stats <- pbapply::pbsapply(cl = cl, simplify = F, experiments, function(experiment) {
    require(data.table)
    source(settings$utility_script)
    message(experiment)
    M <- LFs$maui_factors[[experiment]]
    # get clinical factors for the patients in matrix
    clin <- process_clinical_covariates(rownames(M))
    clin <- get_surv_df(surv, clin[,intersect(colnames(clin), clin_factors),drop = F])
    res <- get_prognostic_covariates(surv, clin, M)
    if(!is.null(res)) {
      cancer_type <- strsplit(experiment, "\\.")[[1]][1]
      res$project <- cancer_type
      v <- res[order(pval)][1]$variable
      dat <- merge(clin, M[,v,drop = F], by = 'row.names')
      rownames(dat) <- dat$Row.names
      dat$Row.names <- NULL
      # compute hazard ratios 
      cox1 <- coxph(Surv(PFI.time,PFI) ~ ., 
              data=dat, x = TRUE)
      hr <- data.frame(summary(cox1)[['conf.int']]) # get hazard ratios with conf intervals
      hr <- hr[grep(v, rownames(hr)), c(1,3,4)]
      colnames(hr) <- c('hr', 'hr_low', 'hr_high')
      rownames(hr) <- v
      p1 <- survminer::ggforest(cox1, data = dat)

@@

      # compute Cindex with all survival-predictive LFs and without LFs 
      vars <- head(res[order(pval)]$variable, 10)
      rsf_stats <- fit_RSF_clin(clin, M[,vars], ntree = 1000)

@@

ggsave(filename = 'surv.Cindex_improv.hazard_ratio.pdf', plot = p, width = 12, height = 14, units = 'in')

Figures <- c(Figures, Fig)
```

@@

p2 <- cowplot::plot_grid(plotlist = plots[1:2], nrow = 2, labels = c('B', 'C'))
Fig <- cowplot::plot_grid(p1, p2, rel_widths = c(3,2), labels = c('A'))
ggsave(filename = 'clinical_factors.LFs.pdf', p, units = 'in', width = 12, height = 9)
Figures <- c(Figures, Fig)
```

@@

# get top 100 features
df <- df[order(abs(df[,top_lf]), decreasing = T),]
freq <- do.call(rbind, lapply(0:29, function(i) {
  s <- (i*100+1)
  e <- (i+1)*100
  x <- data.frame(table(df[s:e,]$data.type))
  x$rank <- stringr::str_wrap(paste0("[",s,",",e,"]"),
                              width = 8)
  x$ind <- i
  return(x)
}))
p3 <- ggplot(freq[freq$ind < 5,], aes(x = reorder(rank, ind), y = Freq)) + 
  geom_bar(aes(fill = Var1), stat='identity') + theme(legend.title = element_blank()) + 
  labs(x = paste0("top features contributing to ",top_lf))

Fig <- cowplot::plot_grid(p1, p3, p2.1, p2.2, labels = 'AUTO')
ggsave(filename = 'MSI.status.pdf', Fig, units = 'in', width = 10, height = 8)

Figures <- c(Figures, Fig)
```

@@

# use hallmark gene sets to annotate cluster specific functions
# TODO: update this to point to /data/local/buyar/arcas/subtyping_paper/data/geneset_scores/singscore
# and remove paths from script (put it in settings.yaml)
scores <- t(readRDS('/data/local/buyar/pancancer_diagnostics/data/geneset_scores/msigdb_hallmarks/TCGA.ssgsea_all.RDS'))
S <- scores[intersect(rownames(M), rownames(scores)),]
clust$tcga <- tcga_subtypes[match(rownames(clust), bcr_patient_barcode)]$Subtype_Selected

@@

# altogether
Fig <-  cowplot::plot_grid(p1, p2, p3, p4, h1$gtable, h2$gtable, 
                         labels = 'AUTO', ncol = 2)
ggsave(filename = paste0('semi_supervised.clustering.', experiment,'.pdf'), 
       Fig, units = 'in', 
       width = 10, height = 12)

Figures <- c(Figures, Fig)
```

@@

# use canonical pathways to annotate prognostic groups
# TODO: update this to point to /data/local/buyar/arcas/subtyping_paper/data/geneset_scores/singscore
# and don't use hard-coded paths 
scores <- t(readRDS('/data/local/buyar/pancancer_diagnostics/data/geneset_scores/msigdb_canonical_pathways/ssgsea.canonical_pathways.RDS'))

@@

# plot examples for which we improve upon tcga
dt <- dcast(stats, experiment ~ source, value.var = 'pval')
dt$improvement <- -log10(dt$maui) - -log10(dt$tcga)
dt <- dt[order(improvement, decreasing = T)]

>>>>

# todo: move this to settings file
inputDir <- '/data/local/buyar/arcas/pancancer_multiomics_manuscript/output'
geneSetScoresFolder <- '/data/local/buyar/arcas/subtyping_paper/data/geneset_scores/singscore'
assay_dir <- settings$assay_output$folder
surv <- do.call(rbind, readRDS(settings$surv))
projectDict <- sapply(yaml::read_yaml(settings$projectDict), 
                      function(x) gsub(' ', '', unlist(strsplit(x, ','))))
tcga_subtypes <- process_tcga_subtypes()

Figures <- list() # accummulate all figures in this
SuppFigures <- list() # accumulate all supplementary figures in this

TOOLS <- c('pca_factors', 'maui_factors', 'mofa_factors', 'mcia_factors')

fig_count <- 0
supFig_count <- 0
```

```{r import_lfs}
LFs <- sapply(simplify = F, TOOLS, function(folder) {
  message('Importing ',folder)
  files <- dir(file.path(inputDir, folder), pattern = 'factors.csv$', full.names = T)
  LFs <- pbapply::pbsapply(simplify = F, files, function(f) {
    dt <- data.table::fread(f)
    M <- as.matrix(dt[,-1])
    rownames(M) <- dt$V1
    return(M)
  })
  names(LFs) <- unlist(lapply(strsplit(basename(names(LFs)), split = '\\.'), function(x) paste(x[1:2], collapse = '.')))
  return(LFs)
})
```

```{r import_feature_weights}
FWs <- sapply(simplify = F, TOOLS, function(folder) {
  message('Importing ',folder)
  files <- dir(file.path(inputDir, folder), pattern = 'feature_weights.csv$', full.names = T)
  W <- pbapply::pbsapply(simplify = F, files, function(f) {
    dt <- data.table::fread(f)
    M <- as.matrix(dt[,-1])
    rownames(M) <- gsub("^assay: ", "", dt$V1) # required only for maui
    return(M)
  })
  names(W) <- unlist(lapply(strsplit(basename(names(W)), split = '\\.'), 
                            function(x) paste(x[1:2], collapse = '.')))
  return(W)
})
```

```{r}
# Define a function that takes as input LFs and some patient labels (clinical/cancer type/subtype)
# and evaluates predictability of the labels using the LFs
predict_labels <- function(LFs, surv, tcga_subtypes, 
                           project, datatype, label.type, ...) {
  source(settings$utility_script)
  predict_stats <- do.call(rbind, sapply(simplify = F, names(LFs), function(tool) {
    require(data.table)
    message("predicting ",label.type," using ",tool)
    experiment <- paste(project, datatype, sep = '.')
    M <- LFs[[tool]][[experiment]]
    if(label.type == 'cancer_type') {
      labels <- data.frame('cancer_type' = surv[match(rownames(M), bcr_patient_barcode)]$project, 
                           row.names = rownames(M))
    } else if (label.type == 'subtype') {
      labels <- data.frame('subtype' = tcga_subtypes[match(rownames(M),
                                                           bcr_patient_barcode)]$Subtype_Selected, 
                           row.names = rownames(M))
    } else if (label.type == 'clinical') {
      labels <- process_clinical_covariates(rownames(M))
    }
    dt <- do.call(rbind, lapply(colnames(labels), function(x) {
      l <- labels[,x]
      if(is.numeric(l)){
        # categorize samples by intervals
        l <- as.character(cut_interval(l, 10))
      }
      l[is.na(l)] <- 'Undefined'
      message("Predicting ",x, " for experiment ",experiment)
      dt <- suppressMessages(predict_cluster_labels(M, l, ...))
      dt$label.type <- x
      dt$project <- project
      dt$datatypes <- datatype
      dt$method <- tool
      return(dt)
    }))
    return(dt)
  }))
  return(predict_stats[!is.na(train_auc)])
}

plot_comp1 <- function(dt) {
  dt$method <- toupper(gsub("_factors", "", dt$method))
  comps <- lapply(setdiff(unique(dt$method), 'MAUI'), function(x) c('MAUI', x))
  ggboxplot(dt, 
          x = 'method', y = 'test_auc', color = 'method', 
          add = 'jitter', legend = 'none') + 
    stat_compare_means(method = 'wilcox.test', paired = T, 
                       method.args = list('alternative' = 'greater'), 
                     comparisons = comps) +
    labs(y = 'AUC') + theme(axis.text = element_text(size = 10))
}

plot_comp2 <- function(dt) {
  dt$method <- toupper(gsub("_factors", "", dt$method))
  dtc <- melt(dcast(dt, label ~ method, value.var = 'test_auc'), 
              id.vars = c('label', 'MAUI'))
  dtc$difference <- dtc$MAUI - dtc$value
  ggplot(dtc, 
         aes(x = MAUI, y = value)) + 
    geom_point() + geom_abline(slope = 1) + 
    facet_wrap(~ variable, nrow = 1) + 
    labs(y = 'AUC')
}

@@

```{r pancancer_prediction, fig.height=9, fig.width=12}
# Find out if cancer types are predictable using latent factors 
pancancer_stats <- predict_labels(LFs, surv, tcga_subtypes, 'pancancer', 
                                  'cnv_gex_meth_mut', label.type = 'cancer_type', num.trees = 1000)
p1 <- plot_comp1(pancancer_stats) + theme(axis.title.x = element_blank())
p2 <- plot_comp2(pancancer_stats)
# plot cancer specific factors
M <- LFs$maui_factors$pancancer.cnv_gex_meth_mut
ct <- surv[match(rownames(M), bcr_patient_barcode)]$project
p3 <- plot_label_specific_vars(M, ct)

@@

# do unsupervised clustering and compute AMI 
experiment <- 'pancancer.cnv_gex_meth_mut'
df <- do.call(rbind, pbapply::pblapply(TOOLS, function(x) {
  message(x)
  M <- LFs[[x]][[experiment]]
  if(x == 'maui_factors') {
    M <- remove_redundant_variables(M, perc = 99)
  }
  # find cancer types per sample
  ct <- surv[match(rownames(M), bcr_patient_barcode)]$project
  data.frame('tool' = x, 
             'AMI' = compute_AMI(M, k = length(unique(ct)), target_labels = ct))
}))
df$tool <- toupper(gsub("_factors", "", df$tool))
p5 <- ggplot(df, aes(x = tool, y = AMI)) +
  geom_bar(aes(fill = tool), stat = 'identity') + 
  geom_text(aes(label = round(AMI, 2))) + 
  theme(legend.title = element_blank(), axis.title.x = element_blank())

Fig <- plot_grid(plot_grid(p1, p2, p5, nrow = 1, labels = c('A', 'B', 'C')), 
                  plot_grid(p3, p4, rel_widths = c(5,3), nrow = 1, labels = c('D', 'E')), 
                  nrow = 2,
                  rel_heights = c(2,3))

fig_count <- fig_count + 1
Figures[[fig_count]] <- Fig
ggsave(paste0('Figure', fig_count, '.pancancer_cancer_type_prediction.pdf'), 
              plot = Fig, units = 'in', width = 12, height = 9)

@@

p1 <- plot_comp1(pancancer_subtype_stats) + theme(axis.title.x = element_blank())
p2 <- plot_comp2(pancancer_subtype_stats)
# plot subtype specific factors
M <- LFs$maui_factors$pancancer.cnv_gex_meth_mut
st <- tcga_subtypes[match(rownames(M), bcr_patient_barcode)]$Subtype_Selected
# remove NA values and pick subtypes with at least 20 samples 
selected <- names(which(table(st) > 20))
M <- M[st %in% selected,]
labels <- tcga_subtypes[match(rownames(M), bcr_patient_barcode)]$Subtype_Selected

@@

# data type contribution to top LF per factor
p4 <- plot_datatype_contribution_by_factor(LFs, FWs, 'maui_factors', 
                                           'pancancer', 'cnv_gex_meth_mut', factors = st, 
                                           top_lfs = 1, top_features = 100) 
p4 <- p4 + labs(y = 'Relative Contribution\nto top latent factor') + 
  theme(legend.position = 'right', legend.background = element_blank())

Fig <- plot_grid(plot_grid(plot_grid(p1, p2 + theme(axis.text.x = element_text(angle = 45)), 
                                     nrow = 2, labels = c('A', 'B')), 
                           p3 + theme(legend.position = 'none', 
                                      axis.text.x = element_text(angle = 45)), 
                           labels = c('', 'C'), ncol = 2, rel_widths = c(1,3)), 
                 p4 + theme(axis.text.x = element_text(angle = 45)), 
                 labels = c('', '', 'D'), ncol = 2, rel_widths = c(2.5,1))

supFig_count <- supFig_count + 1
ggsave(paste0('SuppFigure', supFig_count, '.pancancer_subtype_prediction.pdf'), 
       plot = Fig, units = 'in', width = 18, height = 12)
SuppFigures[[supFig_count]] <- Fig
```

@@

Use top 10 factors (derived from multi-omics) to cluster samples using cancer-specific LFs 
```{r clustering_metrics, fig.width=10, fig.height=8}
multi_omics_comb <- 'cnv_gex_meth_mut'
experiments <- grep(paste0('^TCGA-.*.', multi_omics_comb), names(LFs$pca_factors), value = T)
cl <- parallel::makeCluster(20)
parallel::clusterExport(cl, varlist = c('LFs', 'experiments', 'settings'))
clusters <- sapply(simplify = F, names(LFs), function(tool) {
  message("Clustering top ",tool)
  pbapply::pbsapply(cl = cl, simplify = F, experiments, function(experiment) {
    message(experiment)
    source(settings$utility_script)
    require(data.table)
    M <- LFs[[tool]][[experiment]]
    if(tool == 'maui_factors') {
      M <- remove_redundant_variables(M, perc = 99) # !!! make sure to use same setting as used for pan-cancer clustering
    } 
    M <- t(subset_features_by_variance(t(M), topN = 10))

    val <- c('internal') 
    if(ncol(M) > 2) {val <- c('internal', 'stability')}
    res <- clValid::clValid(obj = M,
                          nClust = 3:8,
                          clMethods = 'kmeans',
                          validation = val,
                          maxitems = 10000,
                          nstart = 1000)
    memberships <- as.data.frame(do.call(cbind, lapply(res@clusterObjs[['kmeans']], function(x) x[['cluster']])))
    colnames(memberships) <- paste0('k', colnames(memberships))
    stats <- as.data.table(res@measures)[V1 %in% c('Silhouette', 'APN')]
    colnames(stats) <- c('metric', 'k', 'clustering_method', 'score')
    return(list('clustering_stats' = stats, 'cluster_memberships' = memberships))
  })
})
parallel::stopCluster(cl)

@@

# plot silhouette/stability score comparison
dt1 <- cluster.stats[metric == 'Silhouette'][datatypes == multi_omics_comb]
dt1$method <- toupper(gsub("_factors", "", dt1$method))
comps <- lapply(setdiff(unique(dt1$method), 'MAUI'), function(x) c('MAUI', x))

p1 <- ggboxplot(dt1, 
          x = 'method', y = 'score', add = 'jitter', color = 'method') + 
  stat_compare_means(method = 'wilcox.test', paired = TRUE, 
                     comparisons = comps) + 
  labs(title = 'Silhouette score', x = '') + 
  theme(legend.position = 'none')

dt2 <- cluster.stats[metric == 'APN'][datatypes == multi_omics_comb]
dt2$method <- toupper(gsub("_factors", "", dt2$method))
comps <- lapply(setdiff(unique(dt2$method), 'MAUI'), function(x) c('MAUI', x))
p2 <- ggboxplot(dt2, 
          x = 'method', y = 'score', add = 'jitter', color = 'method') + 
  stat_compare_means(method = 'wilcox.test', paired = TRUE, 
                     comparisons = comps) + 
  labs(title = 'Cluster Stability Score (APN)', x = '') + 
  theme(legend.position = 'none')

# combine both scores and plot for different values of k
dt <- dcast(cluster.stats[datatypes == multi_omics_comb], 
            cancer_type + method + k ~ metric, value.var = 'score')
dt$kv <- paste0('kmeans(k=',dt$k,')')
dt$method <- toupper(gsub("_factors", "", dt$method))
p3 <- ggplot(dt,
       aes(x = Silhouette, y = APN)) + 
        geom_point(aes(color = method), size = 3) + 
        facet_wrap(~ kv, ncol = 2) +
        labs(title = 'Multi-omics clustering: Silhouette vs APN')

Fig <- plot_grid(plot_grid(p1, p2, labels = 'AUTO', ncol = 1),
                  p3, ncol = 2, rel_widths = c(2,3), labels = c('','C'))

fig_count <- fig_count + 1
Figures[[fig_count]] <- Fig

ggsave(paste0('Figure', fig_count, '.cluster_stability_metrics.pdf'), 
       plot = Fig, units = 'in', 
       width = 10, height = 8)
```

@@

```{r survival_factors, fig.width=12, fig.height=8}
# factors to use in the model
clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage')

experiments <- grep('^TCGA-.*.cnv_gex_meth_mut', names(LFs$maui_factors), value = T)
cl <- parallel::makeCluster(20)
parallel::clusterExport(cl, varlist = c('LFs', 'clin_factors', 
                                        'surv', 'experiments', 'settings'))
surv.stats <- pbapply::pbsapply(cl = cl, simplify = F, experiments, function(experiment) {
    require(data.table)
    source(settings$utility_script)
    message(experiment)
    M <- LFs$maui_factors[[experiment]]
    # get clinical factors for the patients in matrix
    clin <- process_clinical_covariates(rownames(M))
    clin <- subset(clin, select = intersect(colnames(clin), clin_factors))
    res <- get_prognostic_covariates(surv, clin, M, 'PFI', 'PFI.time')
    if(!is.null(res)) {
      cancer_type <- strsplit(experiment, "\\.")[[1]][1]
      res$project <- cancer_type
      v <- res[order(pval)][1]$variable
      dat <- merge(clin, M[,v,drop = F], by = 'row.names')
      rownames(dat) <- dat$Row.names
      dat$Row.names <- NULL
      dat <- get_surv_df(surv, dat) # add PFI values
      # compute hazard ratios 
      cox1 <- coxph(Surv(PFI.time,PFI) ~ ., 
              data=dat, x = TRUE)
      hr <- data.frame(summary(cox1)[['conf.int']]) # get hazard ratios with conf intervals
      hr <- hr[grep(v, rownames(hr)), c(1,3,4)]
      colnames(hr) <- c('hr', 'hr_low', 'hr_high')
      rownames(hr) <- v
      p1 <- survminer::ggforest(cox1, data = dat)

@@

      # compute Cindex with all survival-predictive LFs and without LFs 
      vars <- res[pval < 0.05]$variable
      rsf_stats <- fit_RSF_clin(get_surv_df(surv, clin), M[,vars], ntree = 1000)

@@

fig_count <- fig_count + 1
ggsave(filename = paste0('Figure', fig_count, '.surv.Cindex_improv.hazard_ratio.pdf'), 
                         plot = Fig, width = 12, height = 14, units = 'in')

Figures[[fig_count]] <- Fig
```

@@

p2 <- cowplot::plot_grid(plotlist = plots[1:2], nrow = 2, labels = c('B', 'C'))
Fig <- cowplot::plot_grid(p1, p2, rel_widths = c(3,2), labels = c('A'))
fig_count <- fig_count + 1
ggsave(filename = paste0('Figure', fig_count, '.clinical_factors.LFs.pdf'), 
       plot = Fig, units = 'in', width = 12, height = 9)
Figures[[fig_count]] <- Fig
```

@@

# get top 100 features
df <- df[order(abs(df[,top_lf]), decreasing = T),]
freq <- do.call(rbind, lapply(0:29, function(i) {
  s <- (i*100+1)
  e <- (i+1)*100
  x <- data.frame(table(df[s:e,]$data.type))
  x$rank <- stringr::str_wrap(paste0("[",s,",",e,"]"),
                              width = 8)
  x$ind <- i
  return(x)
}))
p3 <- ggplot(freq[freq$ind < 5,], aes(x = reorder(rank, ind), y = Freq)) + 
  geom_bar(aes(fill = Var1), stat='identity') + 
  theme(legend.title = element_blank(), axis.text.x = element_text(angle = 45)) + 
  labs(x = paste0("top features contributing to ",top_lf))

Fig <- cowplot::plot_grid(p1, 
                          plot_grid(p2.1, p2.2, p3, nrow = 1, labels = c('B', 'C', 'D')),
                          nrow = 2, labels = c('A', ''), rel_heights = c(5,3))
fig_count <- fig_count + 1
ggsave(filename = paste0("Figure", fig_count, '.MSI.status.pdf'), 
                         plot = Fig, units = 'in', width = 10, height = 8)

Figures[[fig_count]] <- Fig
```

@@

# use hallmark gene sets to annotate cluster specific functions
scores <- import_geneset_scores(geneSetScoresFolder, 'msigdb_hallmarks')

S <- scores[intersect(rownames(M), rownames(scores)),]
colnames(S) <- gsub("HALLMARK_", "", colnames(S))
clust$tcga <- tcga_subtypes[match(rownames(clust), bcr_patient_barcode)]$Subtype_Selected

@@

# altogether
Fig <-  cowplot::plot_grid(p1, p2, p3, p4, h1$gtable, h2$gtable, 
                         labels = 'AUTO', ncol = 2)

fig_count <- fig_count + 1
ggsave(filename = paste0('Figure', fig_count, '.semi_supervised.clustering.', experiment,'.pdf'), 
       Fig, units = 'in', 
       width = 10, height = 12)

Figures[[fig_count]] <- Fig
```

@@

# use canonical pathways to annotate prognostic groups
scores <- import_geneset_scores(geneSetScoresFolder, 'msigdb_canonical_pathways')

@@

# plot examples for which we improve upon tcga
dt <- data.table(dcast(stats, experiment ~ source, value.var = 'pval'))
dt$improvement <- -log10(dt$maui) - -log10(dt$tcga)
dt <- dt[order(improvement, decreasing = T)]
